---
title: "data exploration 2"
author: "Andrius consultant group 1"
date: "2025-05-12"
output: html_document
---

```{r}
library(reshape2)
library(dplyr)
library(EBImage)
library(reshape2)
library(umap)
```

# Data loading
```{r}

# Set path to your image folder
image_dir <- "D:/Wanying project repository/Wanying-s-project/raw data/Mediem conc"

# List all .tiff files
tiff_files <- list.files(image_dir, pattern = "\\.tif{1,2}$", full.names = TRUE)

# Read and store images in a named list
marker_images <- lapply(tiff_files, readImage)
names(marker_images) <- tools::file_path_sans_ext(basename(tiff_files))

# Check image dimensions
sapply(marker_images, dim)
marker_images$`ADIPOR-1`
```

As looking at the image we have, I noticed in some images—that the red channel appears completely saturated. This might suggest overexpression of that marker or an issue with image scaling. It would be helpful to revisit the normalization and contrast settings to ensure we’re not losing useful biological information.

# Preprocessing

```{r}
# === Step 0: Load and preprocess DNA channel (Ir191, channel 2) with intensity squaring ===
img <- marker_images[["CD8"]] 
dna <- Image(img[,,2], colormode = "Grayscale")

# Handle NA/NaN/Inf values
dna[is.na(dna)] <- 0
dna[is.infinite(dna)] <- 0

# Square the intensity values to emphasize cells as suggested
# First normalize to 0-1 range
if(max(dna) > min(dna)) {
  dna_norm <- (dna - min(dna)) / (max(dna) - min(dna))
} else {
  dna_norm <- dna * 0
}

# Square the intensity values
dna_squared <- dna_norm^2

# Re-normalize after squaring
dna_squared <- normalize(dna_squared)
dna_squared[is.na(dna_squared)] <- 0
dna_squared[is.infinite(dna_squared)] <- 0

display(dna, method = "raster", title = "original Intensity")


display(dna_squared, method = "raster", title = "Squared Intensity")


# === Step 1: Apply thresholding on squared intensity image ===
# First blur to reduce noise
blurred <- gblur(dna_squared, sigma = 1)
blurred[is.na(blurred)] <- 0

# Thresholding - might need a different offset due to squaring
binary_mask <- thresh(blurred, w = 15, h = 15, offset = 0.15)  # Adjust offset as needed


display(binary_mask, method = "raster", title = "blured mask")



```




# Cell segmentation

```{r}
# === Step 2: Detect bright regions to use as seeds ===
high_intensity <- dna_squared > 0.2  # Tune this threshold if needed
high_intensity <- opening(high_intensity, makeBrush(3, shape = "disc"))  # Remove noise
center_seeds <- bwlabel(high_intensity)

# === Step 3: Segment using propagate ===
segmented <- propagate(x = dna_squared, seeds = center_seeds, mask = binary_mask)

# Clean up
segmented[is.na(segmented)] <- 0
segmented[is.infinite(segmented)] <- 0
segmented <- Image(segmented, colormode = "Grayscale")

# Display segmented result

display(binary_mask, method = "raster", title = "blured mask")


display(colorLabels(segmented), method = "raster", title = "Segmented Cells")
cat("Number of segmented cells:", max(segmented), "\n")


```



```{r}
# Extract red channel (e.g., CD8)
red_marker <- normalize(Image(img[,,1], colormode = "Grayscale"))

# Convert segmentation labels to a normalized grayscale mask
seg_mask <- normalize(segmented > 0)  # binary mask

# Create RGB overlay: red = CD8, green = segmentation mask
overlay_rgb <- rgbImage(
  red = red_marker,
  green = seg_mask,
  blue = red_marker * 0
)

display(overlay_rgb, method = "raster", title = "Segmentation Overlay on CD8 Marker")

```




# Intensity analysis

```{r}
# === Step 0: Clean up and ensure integer-labeled segmentation mask ===
segmented_cells <- round(segmented)
segmented_cells <- as.integer(segmented_cells)

# === Step 1: Get valid cell IDs (exclude background) ===
cell_ids <- sort(unique(segmented_cells))
cell_ids <- cell_ids[cell_ids != 0]  # remove background label 0

# === Step 2: Prepare results matrix ===
marker_names <- names(marker_images)
intensity_matrix <- matrix(NA, nrow = length(cell_ids), ncol = length(marker_names))
rownames(intensity_matrix) <- paste0("Cell_", cell_ids)
colnames(intensity_matrix) <- marker_names

# === Step 3: Loop through each marker image and compute mean intensity per cell ===
for (m in seq_along(marker_names)) {
  marker <- marker_names[m]
  img_stack <- marker_images[[marker]]
  
  # Use channel 1 by default (adjust if needed)
  marker_img <- normalize(Image(img_stack[,,1], colormode = "Grayscale"))
  
  # Handle NA/Inf
  marker_img[is.na(marker_img)] <- 0
  marker_img[is.infinite(marker_img)] <- 0

  # Compute mean intensity per labeled cell
  mean_vals <- tapply(marker_img, segmented_cells, mean)

  # Map results back to the matrix in correct order
  intensity_matrix[, m] <- mean_vals[as.character(cell_ids)]
}

# === Step 4: Convert to data frame for analysis ===
intensity_df <- as.data.frame(intensity_matrix)

intensity_df

```
```{r}
# Convert to long format
intensity_df$CellID <- rownames(intensity_df)
long_df <- reshape2::melt(intensity_df, id.vars = "CellID", variable.name = "ImageID", value.name = "Intensity")

# Clean image names (e.g., remove ".tiff.1")
long_df$ImageID <- sub("\\.tiff\\..*", "", long_df$ImageID)

# Summary stats per image/marker
image_marker_stats <- long_df %>%
  group_by(ImageID) %>%
  summarise(
    b.mean = mean(Intensity, na.rm = TRUE),
    b.sd   = sd(Intensity, na.rm = TRUE),
    b.mad  = mad(Intensity, na.rm = TRUE),
    b.q001 = quantile(Intensity, 0.001, na.rm = TRUE),
    b.q005 = quantile(Intensity, 0.005, na.rm = TRUE),
    b.q05  = quantile(Intensity, 0.05, na.rm = TRUE),
    b.q095 = quantile(Intensity, 0.95, na.rm = TRUE),
    b.q099 = quantile(Intensity, 0.99, na.rm = TRUE)
  )

image_marker_stats
```
This table summarizes the **distribution of marker expression across all cells within each image**. For every marker (represented by the `ImageID`, which corresponds to a specific marker image), it calculates key statistics: the mean intensity (`b.mean`), standard deviation (`b.sd`), and median absolute deviation (`b.mad`), as well as several quantiles (`b.q001`, `b.q005`, `b.q05`, `b.q095`, `b.q099`). These values help capture both the **central tendency and variability** of marker expression across cells. Such summary statistics are useful for identifying **differences in staining intensity**, **signal consistency**, or **outlier markers**, and can aid in **quality control** or **comparing overall marker abundance** between tissue samples.


```{r}
# Step 1: Convert wide to long format
intensity_df$CellID <- rownames(intensity_df)
long_df <- melt(intensity_df, id.vars = "CellID", variable.name = "Marker", value.name = "Intensity")

# Step 2: Create boxplots for each marker, in purple
ggplot(long_df, aes(x = Marker, y = Intensity)) +
  geom_boxplot(fill = "purple", color = "black", outlier.color = "black") +
  theme_minimal() +
  labs(title = "Marker Expression per Cell", x = "Marker", y = "Mean Intensity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r}
# Step 1: Reshape to long format
intensity_df$CellID <- rownames(intensity_df)
long_df <- melt(intensity_df, id.vars = "CellID", variable.name = "Marker", value.name = "Intensity")

# Step 2: Order markers by median intensity (optional but helpful)
marker_order <- long_df %>%
  group_by(Marker) %>%
  summarize(median_int = median(Intensity, na.rm = TRUE)) %>%
  arrange(desc(median_int)) %>%
  pull(Marker)

long_df$Marker <- factor(long_df$Marker, levels = marker_order)

# Step 3: Create violin plot with enhancements
ggplot(long_df, aes(x = Marker, y = Intensity)) +
  geom_violin(fill = "purple", color = "black", alpha = 0.8) +          # purple fill
  geom_boxplot(width = 0.1, outlier.shape = NA, color = "grey80") +     # white boxplot overlay
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +# reference line at 0.5
  theme_minimal(base_size = 12) +
  labs(
    title = "Marker Intensity Distributions",
    subtitle = "Per-cell expression across all segmented cells",
    x = "Marker",
    y = "Mean Intensity"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11)
  )
```












# Clustering



## normalisation

```{r}
# Drop CellID column for clustering input
intensity_mat <- intensity_df[, setdiff(names(intensity_df), "CellID")]

# Optional: log-transform if you have skewed intensities
intensity_log <- log1p(intensity_mat)

# Standardize
scaled_data <- scale(intensity_log)

```


## PCA 
```{r}
pca_res <- prcomp(scaled_data, center = TRUE, scale. = FALSE)
summary(pca_res)  # Check how many PCs explain 80–95% variance

# Extract scores
pc_scores <- as.data.frame(pca_res$x[, 1:5])  
pc_scores$CellID <- intensity_df$CellID

```

## clustering
```{r}
set.seed(1)
kmeans_res <- kmeans(pc_scores[, 1:5], centers = 5)  
pc_scores$Cluster <- factor(kmeans_res$cluster)
umap_res <- umap(pc_scores[, 1:5])
plot(umap_res$layout, col = pc_scores$Cluster, pch = 16,
     main = "UMAP: Cell Clusters by Marker Expression")

```
Here, cells in the same cluster express similar combinations of markers, and are likely to be the same cell type or functional state.

The UMAP projection reveals clear separation between clusters of cells based on their marker expression profiles. This indicates distinct phenotypic subpopulations within the tissue. 


```{r}
# Choose marker to plot
marker_to_plot <- "CD44"

# Extract UMAP layout and matching marker intensity
plot_df <- data.frame(
  UMAP1 = umap_res$layout[, 1],
  UMAP2 = umap_res$layout[, 2],
  Intensity = intensity_df[[marker_to_plot]]
)

# Plot using ggplot2
library(ggplot2)

ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = Intensity)) +
  geom_point(size = 1.5) +
  scale_color_gradient(low = "gray90", high = "red") +
  labs(
    title = paste("UMAP: Cells Colored by", marker_to_plot, "Expression"),
    color = "Intensity"
  ) +
  theme_minimal()

```







