---
title: "Wanying consultation report draft"
author: "Wanying consultant group 2"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
number-sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r library_load,warning = FALSE, message=FALSE}
library(reshape2)
library(dplyr)
library(EBImage)
library(ggrepel)
library(reshape2)
library(umap)
library(e1071)
library(kableExtra)
library(knitr)
library(tibble)
library(DT)
```
# Executive summary

# Introduction & Objextives

The client provided us with imaging mass cytometry (IMC) data from tissue samples and requested assistance with image quantification and cell density analysis. However, recognizing that IMC data offers far more analytical possibilities than basic cell counting, we expanded our analysis to address several key questions that would provide comprehensive insights into the client's samples:

1. Which markers in the panel are performing effectively? Some markers in multiplex imaging panels may not produce reliable signal due to antibody issues, low target expression, or technical limitations. We implemented a systematic marker effectiveness analysis to identify which markers provide meaningful biological information.

2. What are the expression patterns of different markers at the single-cell level? Beyond simple presence/absence, understanding the distribution and heterogeneity of marker expression across cells provides deeper biological insights and helps identify cell subpopulations.

3. How do different markers correlate at the cellular level? Patterns of co-expression can reveal functional relationships between markers and help identify specific cell types or states within the tissue microenvironment.

4. What is the spatial organization of different markers within the tissue? The spatial context of marker expression provides important information about tissue architecture and cellular interactions that is lost in dissociated cell analyses.

By addressing these questions through our analytical pipeline, we aim to provide the client with a comprehensive characterization of their tissue samples that extends well beyond basic cell density measurements, enabling richer biological interpretation and more informed decision-making for their research.




# Data overview & Exploration


We first load in the dataset been provuded:
```{r data_load}
# Here we load in the dataset(images) provided
image_dir <- "raw data/Mediem conc"
tiff_files <- list.files(image_dir, pattern = "\\.tif{1,2}$", full.names = TRUE)
marker_images <- lapply(tiff_files, readImage)
names(marker_images) <- tools::file_path_sans_ext(basename(tiff_files))

```



```{r display, warning=FALSE}
first_marker_name <- names(marker_images)[1] 
first_image <- marker_images[[1]]            

# Create colored versions of the channels
# Extract channels
channel1 <- normalize(first_image[,,1])  
channel2 <- normalize(first_image[,,2])  

# Create RGB images with proper coloring
# For red channel (channel 1)
red_colored <- Image(array(0, dim=c(dim(channel1), 3)), colormode="Color")
red_colored[,,1] <- channel1  
red_colored[,,2] <- 0         
red_colored[,,3] <- 0         

# For cyan channel (channel 2)
cyan_colored <- Image(array(0, dim=c(dim(channel2), 3)), colormode="Color")
cyan_colored[,,1] <- 0        
cyan_colored[,,2] <- channel2 
cyan_colored[,,3] <- channel2  

# Set up side-by-side layout
par(mfrow = c(1, 2), mar = c(0, 0, 0, 0))  

# Plot red channel
display(red_colored, method = "raster")
text(x = 10, y = 20, labels = "Figure 1: Marker Channel (Red)", col = "white", cex = 1, pos = 4)

# Plot cyan channel
display(cyan_colored, method = "raster")
text(x = 10, y = 20, labels = "Figure 2: Nuclei Channel (Cyan)", col = "white", cex = 1, pos = 4)

# Reset layout
par(mfrow = c(1, 1))
```



The *Figure 1* and *Figure 2* above displays the two channels of our imaging mass cytometry data for the marker ADIPOR-1 (the first marker in our dataset). The left panel shows Channel 1 (red), which contains the specific marker signal that indicates protein expression. The right panel shows Channel 2 (cyan), which represents the DNA/nuclear signal that delineates individual cells. These two channels provide complementary information: Channel 1 allows us to assess the distribution and intensity of specific protein markers across the tissue sample, while Channel 2 enables identification of cellular structures and can be used for cell segmentation. 





# Analysis

## 1. Marker Effectiveness Analysis

**Mean Intensity**
- Formula: $\mu = \frac{1}{n}\sum_{i=1}^{n}x_i$
- Represents the average signal intensity across all pixels in the marker channel
- Higher values generally indicate stronger marker expression

**Median Intensity**
- Formula: Middle value of all pixel intensities when sorted
- Represents the typical intensity value, less affected by outliers than mean
- Zero median suggests most pixels have no signal

**Standard Deviation (SD)**
- Formula: $\sigma = \sqrt{\frac{1}{n}\sum_{i=1}^{n}(x_i-\mu)^2}$
- Measures the spread of intensity values around the mean

**Coefficient of Variation (CV)**
- Formula: $CV = \frac{\sigma}{\mu}$
- Normalizes standard deviation by the mean
- Indicates signal variability relative to signal strength
- High values suggest noisy signal or high heterogeneity

**Skewness**
- Formula: $\text{Skewness} = \frac{1}{n}\sum_{i=1}^{n}(\frac{x_i-\mu}{\sigma})^3$
- Measures asymmetry of the intensity distribution
- Positive values indicate longer tail toward higher intensities
- Higher values typically suggest clearer positive cell populations

**Entropy**
- Formula: $\text{Entropy} = -\sum_{i} p_i \log_2(p_i)$
- Where $p_i$ is the probability of pixel intensity falling in bin $i$
- Quantifies randomness or organization of marker distribution
- Lower values indicate more structured, non-random distribution (typically better)

**Quality Score**
- Formula: $\text{Quality Score} = Z(\text{Mean}) + Z(\text{Skewness}) - Z(\text{Entropy})$
- Where $Z()$ indicates standardization: $Z(x) = \frac{x - \mu_x}{\sigma_x}$
- Combined score that rewards high mean intensity, high skewness, and low entropy
- Used for ranking markers from most to least informative


```{r}
#############################################################
## Marker Effectiveness Analysis
#############################################################

# Calculate statistics for all markers
marker_stats <- data.frame()

for (marker_name in names(marker_images)) {
  # Extract marker channel from image
  img <- marker_images[[marker_name]]
  marker_channel <- img[,,1]  # Channel 1 (marker)
  
  # Get values and calculate statistics
  marker_values <- as.vector(marker_channel)
  
  # Basic statistics
  mean_val <- mean(marker_values, na.rm = TRUE)
  median_val <- median(marker_values, na.rm = TRUE)
  sd_val <- sd(marker_values, na.rm = TRUE)
  cv <- sd_val / mean_val
  
  # SNR calculation
  top_10 <- quantile(marker_values, 0.9, na.rm = TRUE)
  bottom_10 <- quantile(marker_values, 0.1, na.rm = TRUE)
  signal <- mean(marker_values[marker_values >= top_10], na.rm = TRUE)
  noise <- mean(marker_values[marker_values <= bottom_10], na.rm = TRUE)
  
  # Skewness and entropy
  skew <- skewness(marker_values, na.rm = TRUE)
  
  # Calculate entropy
  breaks <- seq(min(marker_values, na.rm = TRUE), max(marker_values, na.rm = TRUE), length.out = 20)
  bins <- cut(marker_values, breaks = breaks)
  bin_counts <- table(bins)
  bin_probs <- bin_counts / sum(bin_counts)
  entropy <- -sum(bin_probs * log2(bin_probs + 1e-10))
  
  # Add to results
  marker_stats <- rbind(marker_stats, data.frame(
    Marker = marker_name,
    Mean = mean_val,
    Median = median_val,
    SD = sd_val,
    CV = cv,
    Skewness = skew,
    Entropy = entropy
  ))
}

# Create quality score and rank markers
marker_stats$QualityScore <- scale(marker_stats$Mean) + 
                            scale(marker_stats$Skewness) - 
                            scale(marker_stats$Entropy)

# Sort by quality score
marker_stats <- marker_stats[order(marker_stats$QualityScore, decreasing = TRUE), ]

# Identify non-informative markers
threshold <- quantile(marker_stats$QualityScore, 0.25, na.rm = TRUE)
non_informative <- marker_stats$Marker[marker_stats$QualityScore <= threshold]

# Display results
print(marker_stats[, c("Marker", "Mean", "Median", "CV", "Skewness", "Entropy", "QualityScore")])
cat("Non-informative markers:", paste(non_informative, collapse = ", "), "\n")


```

```{r histogram}
# Create visualization with smaller y-axis text
quality_plot <- ggplot(marker_stats, aes(x = reorder(Marker, QualityScore), y = QualityScore)) +
  geom_bar(stat = "identity", aes(fill = QualityScore)) +
  scale_fill_gradient(low = "red", high = "green") +
  coord_flip() +
  labs(title = "Marker Quality Assessment", x = "Marker", y = "Quality Score") +
  theme_minimal() +
  # Add these lines to reduce y-axis text size
  theme(
    axis.text.y = element_text(size = 7),    # Smaller text for marker names
    axis.title = element_text(size = 9)     # Slightly smaller axis titles
  )

print(quality_plot)
```

```{r buble_plot,fig.width=6, fig.height=5}
bubble_plot <- ggplot(marker_stats, aes(x = Skewness, y = CV)) +
  geom_point(aes(size = Mean, color = Entropy), alpha = 0.7) +
  scale_color_gradient(low = "green", high = "red") +
  # Replace geom_text with geom_text_repel
  geom_text_repel(aes(label = Marker), size = 2, 
                 box.padding = 0.5,
                 point.padding = 0.2,
                 force = 5,
                 segment.size = 0.2,
                 max.overlaps = 30) +
  labs(title = "Marker Quality Bubble Plot",
       x = "Skewness (higher is better)",
       y = "Coefficient of Variation",
       size = "Mean Intensity",
       color = "Entropy (lower is better)") +
  theme_minimal() +
  theme(legend.position = "right")

print(bubble_plot)
```


```{r}
# Display best and worst markers
best_marker <- marker_stats$Marker[1]
worst_marker <- marker_stats$Marker[nrow(marker_stats)]

par(mfrow = c(1, 2))
display(normalize(marker_images[[best_marker]][,,1]), 
        method = "raster", title = paste("Best:", best_marker))
display(normalize(marker_images[[worst_marker]][,,1]), 
        method = "raster", title = paste("Worst:", worst_marker))
par(mfrow = c(1, 1))
```
Here is the comparison of the good marker and a useless marker, ss looking at the image we have in the right hand side, we noticed in some images—that the red channel appears completely saturated. This might suggest overexpression of that marker or an issue with image scaling. It would be helpful to revisit the normalization and contrast settings to ensure we’re not losing useful biological information.Thus we decided to remove those images(markers) in our further analysis.

We identified and removed the least informative markers (those in the bottom 25% of quality scores). These markers showed characteristics of poor signal quality, including low mean intensity, low skewness, and/or high entropy, suggesting they may not reliably distinguish positive from negative cells. By filtering out these non-informative markers, we can focus our subsequent analyses on the most reliable markers, improving the robustness of our cell segmentation and clustering results. The next step in our analysis pipeline is cell segmentation using the DNA channel (Channel 2), followed by quantification of marker expression at the single-cell level using only the informative markers we've identified.


```{r}
# Create filtered marker list
informative_markers <- setdiff(names(marker_images), non_informative)
marker_images_filtered <- marker_images[informative_markers]

# Check how many markers we retained
cat("Retained", length(marker_images_filtered), "informative markers out of", 
    length(marker_images), "total markers\n")

```

## 2. Cell Segmentation

To enable single-cell analysis of immune marker expression across mouse liver tissue, we performed cell segmentation using the **nuclei channel** as a reference. This step is essential for quantifying marker intensities at the cellular level and for linking spatial patterns to immune phenotypes.


### 2.1 Segmentation

We already filtered out non-informative markers based on statistical properties such as **mean intensity**, **coefficient of variation**, **skewness**, and **entropy**—retaining only those with strong and biologically interpretable signals. Using the most robust nuclei-related marker as the segmentation reference, we extracted the **second channel** (commonly *Ir191*) and enhanced nuclear contrast by **squaring normalised intensity values**.

Following **Gaussian smoothing** and **adaptive thresholding**, we generated a **binary mask** to define nuclear regions and then applied a **seed-based watershed algorithm** (`propagate`) to segment individual cells. This pipeline resulted in accurate delineation of cell boundaries across the tissue section, forming a reliable basis for downstream **clustering** and **spatial immune profiling**.


```{r}
#############################################################
## Cell Segmentation Using Reference DNA Channel
#############################################################

# Use the first filtered marker as reference for DNA channel
reference_marker <- names(marker_images_filtered)[1]

# === Step 0: Load and preprocess DNA channel (Ir191, channel 2) with intensity squaring ===
img <- marker_images_filtered[[reference_marker]] 
dna <- Image(img[,,2], colormode = "Grayscale")

# Handle NA/NaN/Inf values
dna[is.na(dna)] <- 0
dna[is.infinite(dna)] <- 0

# Square the intensity values to emphasize cells as suggested
# First normalize to 0-1 range
if(max(dna) > min(dna)) {
  dna_norm <- (dna - min(dna)) / (max(dna) - min(dna))
} else {
  dna_norm <- dna * 0
}

# Square the intensity values
dna_squared <- dna_norm^2

# Re-normalize after squaring
dna_squared <- normalize(dna_squared)
dna_squared[is.na(dna_squared)] <- 0
dna_squared[is.infinite(dna_squared)] <- 0


# === Step 1: Apply thresholding on squared intensity image ===
# First blur to reduce noise
blurred <- gblur(dna_squared, sigma = 1)
blurred[is.na(blurred)] <- 0

# Thresholding - might need a different offset due to squaring
binary_mask <- thresh(blurred, w = 15, h = 15, offset = 0.15)  # Adjust offset as needed

# === Step 2: Detect bright regions to use as seeds ===
high_intensity <- dna_squared > 0.2  # Tune this threshold if needed
high_intensity <- opening(high_intensity, makeBrush(3, shape = "disc"))  # Remove noise
center_seeds <- bwlabel(high_intensity)

# === Step 3: Segment using propagate ===
segmented <- propagate(x = dna_squared, seeds = center_seeds, mask = binary_mask)

# Clean up
segmented[is.na(segmented)] <- 0
segmented[is.infinite(segmented)] <- 0
segmented <- Image(segmented, colormode = "Grayscale")

# Create overlay visualizations for filtered markers
marker_overlays <- list()

for (marker_name in names(marker_images_filtered)) {
  # Extract red channel (marker channel)
  marker_img <- marker_images_filtered[[marker_name]]
  red_marker <- normalize(Image(marker_img[,,1], colormode = "Grayscale"))
  
  # Convert segmentation labels to a normalized grayscale mask
  seg_mask <- normalize(segmented > 0)  # binary mask
  
  # Create RGB overlay: red = marker, green = segmentation mask
  overlay_rgb <- rgbImage(
    red = red_marker,
    green = seg_mask,
    blue = red_marker * 0
  )
  
  marker_overlays[[marker_name]] <- overlay_rgb
}


# 2x2 Panel of segmentation steps
par(mfrow = c(2, 2), mar = c(2, 2, 3, 2))  # Adjust margin for titles

# Plot 1: Original DNA channel
display(dna, method = "raster")
text(x = 10, y = 20, labels = "Figure 3: Original DNA Channel", col = "white", cex = 1.5, pos = 4)

# Plot 2: Squared intensity image
display(dna_squared, method = "raster")
text(x = 10, y = 20, labels = "Figure 4: Squared Intensity", col = "white", cex = 1.5, pos = 4)

# Plot 3: Binary mask after thresholding
display(binary_mask, method = "raster")
text(x = 10, y = 20, labels = "Figure 5: Thresholded Binary Mask", col = "white", cex = 1.5, pos = 4)

# Plot 4: Final segmented cell labels
display(colorLabels(segmented), method = "raster")
text(x = 10, y = 20, labels = "Figure 6: Segmented Cells", col = "white", cex = 1.5, pos = 4)

# Reset layout
par(mfrow = c(1, 1))
```

This panel illustrates the key steps in our cell segmentation workflow. **Figure 3** shows the original grayscale image of the nuclei channel, where individual nuclei are visible but slightly diffuse. To enhance contrast, the normalized intensity was squared, as seen in **Figure 4**, making nuclear regions more distinguishable from the background. After Gaussian blurring and adaptive thresholding, we generated a binary mask (**Figure 5**) that highlights high-intensity nuclear regions likely to correspond to cells. Finally, **Figure 6** displays the segmented cells using a watershed-like algorithm (`propagate`), with each cell region shown in a unique color. This segmentation forms the basis for per-cell marker quantification and downstream analysis.



```{r}
#############################################################
## Step 2: Display Overlay of First Marker with Segmentation
#############################################################

# Display overlay for the first marker with segmentation mask
display(marker_overlays[[1]], method = "raster")
text(x = 10, y = 20,
     labels = paste("Figrue 7: Overlay with Marker -", names(marker_overlays)[1]),
     col = "white", cex = 1.5, pos = 4)


```
This figure shows an overlay of the B220 marker signal (**red**) with the segmented nuclei regions (**green**) as an example. The red channel highlights the spatial expression pattern of B220 across the liver tissue, while the green segmentation mask outlines individual nuclei identified during the segmentation process. The strong co-localization between signal and segmentation indicates that the segmentation pipeline accurately captures nuclear structures, enabling reliable per-cell marker quantification.

### 2.2 Intensity analsyis

To quantify marker expression at the single-cell level, we calculated the mean intensity of each marker within segmented cell regions. The following tables summarize the intensity matrix and key distribution statistics for each marker image.

```{r}
#############################################################
## Per-Cell Marker Intensity Analysis
#############################################################

# === Step 0: Clean up and ensure integer-labeled segmentation mask ===
segmented_cells <- round(segmented)
segmented_cells <- as.integer(segmented_cells)

# === Step 1: Get valid cell IDs (exclude background) ===
cell_ids <- sort(unique(segmented_cells))
cell_ids <- cell_ids[cell_ids != 0]  # remove background label 0

# === Step 2: Prepare results matrix ===
marker_names <- names(marker_images_filtered)
intensity_matrix <- matrix(NA, nrow = length(cell_ids), ncol = length(marker_names))
rownames(intensity_matrix) <- paste0("Cell_", cell_ids)
colnames(intensity_matrix) <- marker_names

# === Step 3: Loop through each marker image and compute mean intensity per cell ===
for (m in seq_along(marker_names)) {
  marker <- marker_names[m]
  img_stack <- marker_images_filtered[[marker]]
  
  # Use channel 1 by default (adjust if needed)
  marker_img <- normalize(Image(img_stack[,,1], colormode = "Grayscale"))
  
  # Handle NA/Inf
  marker_img[is.na(marker_img)] <- 0
  marker_img[is.infinite(marker_img)] <- 0

  # Compute mean intensity per labeled cell
  mean_vals <- tapply(marker_img, segmented_cells, mean)

  # Map results back to the matrix in correct order
  intensity_matrix[, m] <- mean_vals[as.character(cell_ids)]
}

intensity_df <- as.data.frame(intensity_matrix)

# Start with the full intensity data frame
intensity_preview <- intensity_df

# Convert rownames to a column
intensity_preview <- rownames_to_column(intensity_preview, var = "Cell ID")

# Round all numeric columns to 3 decimal places (excluding "Cell ID")
intensity_preview <- intensity_preview %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

# Display as interactive datatable
datatable(
  intensity_preview,
  caption = "Table 2. Per-Cell Marker Intensity Matrix",
  rownames = FALSE,
  options = list(
    pageLength = 6,
    autoWidth = TRUE,
    scrollX = TRUE
  )
)

```

The *Table 2* displays the average expression intensity of immune markers for the first six segmented cells. Each **row** corresponds to a unique cell, while each **column** represents a specific marker. The values reflect the **mean pixel intensity** of each marker within the segmented area of each cell, after normalization and cleaning. These values serve as input for downstream analyses such as clustering, thresholding, and cell type classification based on marker expression profiles. Scroll functionality is provided to explore all 567 cells and the full set of markers.

```{r}
# Convert to long format
intensity_df$CellID <- rownames(intensity_df)
long_df <- reshape2::melt(intensity_df, id.vars = "CellID", variable.name = "ImageID", value.name = "Intensity")

# Clean image names (e.g., remove ".tiff.1")
long_df$ImageID <- sub("\\.tiff\\..*", "", long_df$ImageID)

# Summary stats per image/marker
image_marker_stats <- long_df %>%
  group_by(ImageID) %>%
  summarise(
    b.mean = mean(Intensity, na.rm = TRUE),
    b.sd   = sd(Intensity, na.rm = TRUE),
    b.mad  = mad(Intensity, na.rm = TRUE),
    b.q001 = quantile(Intensity, 0.001, na.rm = TRUE),
    b.q005 = quantile(Intensity, 0.005, na.rm = TRUE),
    b.q05  = quantile(Intensity, 0.05, na.rm = TRUE),
    b.q095 = quantile(Intensity, 0.95, na.rm = TRUE),
    b.q099 = quantile(Intensity, 0.99, na.rm = TRUE)
  )

# Round all numeric summary stats to 3 decimal places
image_marker_stats_rounded <- image_marker_stats %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

# Display in an interactive DT table
datatable(
  image_marker_stats_rounded,
  caption = "Table 3. Summary Statistics for Each Marker Image",
  rownames = FALSE,
  options = list(
    pageLength = 10,
    autoWidth = TRUE,
    scrollX = TRUE
  )
)
```
Next, this *Table 3* summarises the distribution of marker expression across all cells within each image. For every marker (represented by the ImageID, which corresponds to a specific marker image), it calculates key statistics: the mean intensity (b.mean), standard deviation (b.sd), and median absolute deviation (b.mad), as well as several quantiles (b.q001, b.q005, b.q05, b.q095, b.q099). These values help capture both the central tendency and variability of marker expression across cells. Such summary statistics are useful for identifying differences in staining intensity, signal consistency, or outlier markers, and can aid in quality control or comparing overall marker abundance between tissue samples.

```{r}
# Step 1: Reshape to long format
intensity_df$CellID <- rownames(intensity_df)
long_df <- reshape2::melt(intensity_df, id.vars = "CellID", variable.name = "Marker", value.name = "Intensity")

# Step 2: Order markers by median intensity (optional but helpful)
marker_order <- long_df %>%
  group_by(Marker) %>%
  summarize(median_int = median(Intensity, na.rm = TRUE)) %>%
  arrange(desc(median_int)) %>%
  pull(Marker)

long_df$Marker <- factor(long_df$Marker, levels = marker_order)

# Step 3: Create violin plot with enhancements
ggplot(long_df, aes(x = Marker, y = Intensity)) +
  geom_violin(fill = "purple", color = "black", alpha = 0.8) +          # purple fill
  geom_boxplot(width = 0.1, outlier.shape = NA, color = "grey80") +     # white boxplot overlay
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") + # reference line at 0.5
  theme_minimal(base_size = 12) +
  labs(
    title = "Figure 8: Marker Intensity Distributions",
    subtitle = "Per-cell expression across all segmented cells",
    x = "Marker",
    y = "Mean Intensity"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11)
  )
```
Finally, this violin plot (*Figure 8*) shows the distribution of mean intensity values for each marker across all segmented cells. Each violin represents the density and spread of per-cell expression levels, with wider sections indicating higher cell frequency at a given intensity. The horizontal dashed line at 0.5 helps visually compare markers with higher or lower expression levels. Overall, the plot provides an overview of marker variability, aiding in identifying highly expressed markers and assessing their potential relevance for downstream analysis.

## 3. Marker relationships & Co-expression Analysis











## 4. Cell Subpopulation Analysis – Clustering 



# Conclusion

# Limitations

# Appendix